The Third Bellman Equation

### 2019-01-29 paper/thirdbell.tex

A stochastic version can be
\begin{equation}
Z\gets (1+\eta) Z-Z\tilde \Delta Z
\end{equation}
with a stochastic estimate $\tilde \Delta$ of $\Delta$ (obtained from a
transition observed in the MDP). Usually in RL we can only sample
transitions $s\to s'$ from some initial state $s$ distributed wrt some
distribution $\rho$ (which we may not control). This means we can sample
from $\rho \Delta$, more precisely,
\begin{equation}
\rho\Delta=\E_{s\sim \rho,\,s'\sim P(s'|s)}
\1_s\transp{(\1_s-\gamma\1_{s'})}
\end{equation}
A solution is to parameterize
\begin{equation}
Z_{s_1s_2}\eqd \,\E_{s_2\sim \rho} z(s_1,s_2)
\end{equation}
namely, $z$ is the density of $Z$ with respect to $\rho(s_2)$.
Equivalently, in the matrix case, $Z=z\rho$ with $\rho$ the diagonal
matrix with entries $\rho$. Then the Newton update becomes
\begin{equation}
z\rho\gets 2z\rho-z\rho\Delta z\rho
\end{equation}
which is equivalent to
\begin{equation}
z\gets 2z-z\rho\Delta z
\end{equation}
and now $\rho \Delta$ is samplable,
so we can apply the stochastic version of the Newton update. Explicitly
this is
\begin{equation}
z(s_1,s_2)\gets (1+\eta) z(s_1,s_2)-
\E_{s\sim \rho,\,s'\sim P(s'|s)}z(s_1,s)\left(z(s,s_2)-\gamma
z(s',s_2)\right)
\end{equation}

@@

In the non-tabular case, we parameterize $z(s_1,s_2)$ by a machine
learning model. Then, given a current value $z_0$ of $z$, we to TD-like
steps towards the Newton update. More precisely, we define
\begin{equation}
\ztar\deq 2z_0-z_0\rho\Delta z_0
\end{equation}
and perform any number of SGD steps on the parameters of $z$ to reduce
the error $\E_{s_1\sim \rho,\,s_2\sim \rho}
\,(z(s_1,s_2)-\ztar(s_1,s_2))^2$, namely, we draw two states $s_1$ and
$s_2$ according to $\rho$, a transition $s\to s'$ with $s\sim \rho$, and
updated the parameter $\theta$ of $z$
by
\begin{equation}
\theta\gets \theta+\eta \partial_\theta z(s_1,s_2)\left(
2z_0(s_1,s_2)-z_0(s_1,s)\left(z_0(s,s_2)-\gamma z_0(s',s_2)\right)-z(s_1,s_2)
\right)
\end{equation}
which is an unbiased estimate of the gradient of the error.

@@

\paragraph{Parameterizing $Z-\Id$ in the continuous case.} I think it is quite
important to initialize at $Z=\Id$, but this is not easy in the
continuous case (initializing a neural network $z(s_1,s_2)$ to
$\delta_{s_1s_2}/\rho(s_2)$ is not convenient). Moreover, there is an
analytical singularity: for a discrete-time Markov process in continuous
space, the fundamental matrix $Z=\sum_t \gamma^t P^t$ does have two
analytically distinct components: for every $t\geq 1$ the term $P^t$ is
smooth, while for $t=0$ it is a Dirac component along the diagonal
$s_1=s_2$ (corresponding to $\Id$). So it may not be a good idea to
represent them together in the same network. Instead, we can choose a
representation of the operator $Z$ as $Z=\Id+\E_\rho \zeta$ or more
precisely
\begin{equation}
Z_{s_1s_2}=\delta_{s_1s_2}+\E_{s_2\sim \rho} \zeta(s_1,s_2)
\end{equation}
so that the operator $Z$ acting on a reward $R$ is
\begin{equation}
V(s_1)=(ZR)(s_1)=R(s_1)+\int_{s_2} \zeta(s_1,s_2)R(s_2)\d\rho(s_2)
\end{equation}

@@

The non-samplability problem solves itself when considering the gradient of the
error. Let $\zetatar$ be the target above on $\zeta$. Define the loss
$\E_{s_1\sim \rho,\,s_2\sim \rho} (\zeta(s_1,s_2)-\zetatar(s_1,s_2))^2$
as above, whose gradient is
\begin{equation}
\E_{s_1\sim \rho,\,s_2\sim \rho} \,\partial \zeta(s_1,s_2)(\zeta(s_1,s_2)-\zetatar(s_1,s_2))
\end{equation}
and let us study the contributions from each term of $\zetatar$, starting
with the non-samplable term $\gamma P \rho^{-1}$. Its contribution to the
gradient is
\begin{align}
\E_{s_1\sim \rho,\,s_2\sim \rho} \,\partial \zeta(s_1,s_2)
P_{s_1s_2}/\rho(s_2)&=\E_{s_1\sim \rho}\sum_{s_2} \partial
\zeta(s_1,s_2)P_{s_1s_2}
\\&=\E_{s_1\sim \rho,\,s_2\sim P(s_2|s_1)}\partial
\zeta(s_1,s_2)
\end{align}
that is, the gradient of $\zeta$ along a transition $s_1\to s_2$.
So each time we sample a transition $s\to s'$, the term $
P\rho^{-1}$ contributes $\gamma \partial
\zeta(s,s')$ to the gradient. (This differs from the updates above, in
that above we sampled $s\to s'$ and then modified the value of $z$ at
$s_1$ and $s_2$ for independently sampled states $s_1$ and $s_2$; here
sampling $s\to s'$ contributes a change of $\zeta$ at $s$ and $s'$.)

Likewise, the gradient contribution of the $P\zeta$ term is
\begin{align}
\E_{s_1\sim \rho,\,s_2\sim \rho} \,\partial \zeta(s_1,s_2)
\,(P\zeta)(s_1,s_2)&=\E_{s_1\sim \rho,\,s_2\sim \rho}
\partial \zeta(s_1,s_2)\E_{s_3\sim P(s_3|s_1)}\zeta(s_3,s_2)
\end{align}
so this time the transition is $s_1\to s_3$. That is, when we sample a
transition $s\to s'$, we have a gradient term $\E_{s_2\sim \rho}
\partial\zeta(s,s_2)\zeta(s',s_2)$.

@@

Thus, when sampling $s\to s'$, we have several gradient contributions:
one is similar to what we have before and involves
$\partial\zeta(s_1,s_2)$ for independent samples $s_1,s_2$; one involves
$\partial\zeta(s,s')$, one involves $\partial\zeta(s,s_2)$, and the last
involves $\partial\zeta(s_1,s')$. All are samplable.

TODO rewrite gradient with current $\zeta$ and previous $\zeta_0$, with
all correct signs and $\gamma$ factors...

>>>>

A stochastic version can be
\begin{equation}
Z\gets (1+\eta) Z-Z\tilde \Delta Z
\end{equation}
with a stochastic estimate $\tilde \Delta$ of $\Delta$ (obtained from a
transition observed in the MDP). Usually in RL we can only sample
transitions $s\to s'$ from some initial state $s$ distributed wrt some
distribution $\rho$ (which we may not control). This means we can sample
from $\rho \Delta$, more precisely,
\begin{equation}
\rho\Delta=\E_{s\sim \rho,\,s'\sim P(s'|s)}
\1_s\transp{(\1_s-\gamma\1_{s'})}
\end{equation}
A solution is to parameterize
\begin{equation}
Z_{s_1s_2}\eqd \,\E_{s_2\sim \rho} z(s_1,s_2)
\end{equation}
namely, $z$ is the density of $Z$ with respect to $\rho(s_2)$ (and as
matrices,
$Z=z\rho$).
Equivalently, in the matrix case, $Z=z\rho$ with $\rho$ the diagonal
matrix with entries $\rho$. Then the Newton update becomes
\begin{equation}
z\rho\gets 2z\rho-z\rho\Delta z\rho
\end{equation}
which is equivalent to
\begin{equation}
z\gets 2z-z\rho\Delta z
\end{equation}
and now $\rho \Delta$ is samplable,
so we can apply the stochastic version of the Newton update. Explicitly
this is
\begin{equation}
z(s_1,s_2)\gets (1+\eta) z(s_1,s_2)-
\E_{s\sim \rho,\,s'\sim P(s'|s)}z(s_1,s)\left(z(s,s_2)-\gamma
z(s',s_2)\right)
\end{equation}

@@

In the non-tabular case, we parameterize $z(s_1,s_2)$ by a machine
learning model. Then, given a current value $z_0$ of $z$, we to TD-like
steps towards the Newton update. More precisely, we define
\begin{equation}
\ztar\deq 2z_0-z_0\rho\Delta z_0
\end{equation}
and perform any number of SGD steps on the parameters of $z$ to reduce
the error 
\begin{equation}
\ell(z)\deq\tfrac12\, \E_{s_1\sim \rho,\,s_2\sim \rho}
\,(z(s_1,s_2)-\ztar(s_1,s_2))^2
\end{equation}
namely, we draw two states $s_1$ and
$s_2$ according to $\rho$, a transition $s\to s'$ with $s\sim \rho$, and
update the parameter $\theta$ of $z$
by the gradient
\begin{equation}
\partial\ell =-\partial_\theta z(s_1,s_2)\left(
2z_0(s_1,s_2)-z_0(s_1,s)\left(z_0(s,s_2)-\gamma z_0(s',s_2)\right)-z(s_1,s_2)
\right)
\end{equation}
which is an unbiased estimate of the gradient $\partial \ell$ of the error.

@@

\paragraph{Parameterizing $Z-\Id$ in the continuous case.} I think it is quite
important to initialize at $Z=\Id$, but this is not easy in the
continuous case (initializing a neural network $z(s_1,s_2)$ to
$\delta_{s_1s_2}/\rho(s_2)$ is not convenient). Moreover, there is an
analytical singularity: for a discrete-time Markov process in continuous
space, the fundamental matrix $Z=\sum_t \gamma^t P^t$ does have two
analytically distinct components: for every $t\geq 1$ the term $P^t$ is
smooth, while for $t=0$ it is a Dirac component along the diagonal
$s_1=s_2$ (corresponding to $\Id$). So it may not be a good idea to
represent them together in the same network. Instead, we can choose a
representation of the operator $Z$ as $Z=\Id+\zeta\rho$ or equivalently
\begin{equation}
Z_{s_1s_2}=\delta_{s_1s_2}+\E_{s_2\sim \rho} \zeta(s_1,s_2)
\end{equation}
so that the operator $Z$ acting on a reward $R$ is
\begin{equation}
V(s_1)=(ZR)(s_1)=R(s_1)+\int_{s_2} \zeta(s_1,s_2)R(s_2)\d\rho(s_2)
\end{equation}

@@

The non-samplability problem solves itself when considering the gradient of the
error. Let $\zetatar$ be the target above on $\zeta$. Define the loss
$\E_{s_1\sim \rho,\,s_2\sim \rho} (\zeta(s_1,s_2)-\zetatar(s_1,s_2))^2$
as above, whose gradient is
\begin{equation}
\E_{s_1\sim \rho,\,s_2\sim \rho} \,\partial \zeta(s_1,s_2)(\zeta(s_1,s_2)-\zetatar(s_1,s_2))
\end{equation}
and let us study the contributions from each term of $\zetatar$, starting
with the non-samplable term $\gamma P \rho^{-1}$. Its contribution to the
gradient is
\begin{align}
\E_{s_1\sim \rho,\,s_2\sim \rho} \,\partial \zeta(s_1,s_2)
P_{s_1s_2}/\rho(s_2)&=\E_{s_1\sim \rho}\sum_{s_2} \partial
\zeta(s_1,s_2)P_{s_1s_2}
\\&=\E_{s_1\sim \rho,\,s_2\sim P(s_2|s_1)}\partial
\zeta(s_1,s_2)
\end{align}
that is, the gradient of $\zeta$ along a transition $s_1\to s_2$.
So each time we sample a transition $s\to s'$, the term $
\gamma P\rho^{-1}$ contributes $\gamma \partial
\zeta(s,s')$ to the gradient. (This differs from the updates above, in
that above we sampled $s\to s'$ and then modified the value of $z$ at
$s_1$ and $s_2$ for independently sampled states $s_1$ and $s_2$; here
sampling $s\to s'$ contributes a change of $\zeta$ at $s$ and $s'$.)

Likewise, the gradient contribution of the $P\zeta$ term is
\begin{align}
\E_{s_1\sim \rho,\,s_2\sim \rho} \,\partial \zeta(s_1,s_2)
\,(P\zeta)(s_1,s_2)&=\E_{s_1\sim \rho,\,s_2\sim \rho}
\partial \zeta(s_1,s_2)\E_{s_3\sim P(s_3|s_1)}\zeta(s_3,s_2)
\end{align}
so this time the transition is $s_1\to s_3$. That is, when we sample a
transition $s\to s'$, we have a gradient term $\E_{s_2\sim \rho}
\partial\zeta(s,s_2)\zeta(s',s_2)$. This term is absolutely similar to
standard TD (for which $s_2$ is fixed).

@@

Thus, when sampling $s\to s'$, we have several gradient contributions:
one is similar to what we have before and involves
$\partial\zeta(s_1,s_2)$ for independent samples $s_1,s_2$; one involves
$\partial\zeta(s,s')$, one involves $\partial\zeta(s,s_2)$, and the last
involves $\partial\zeta(s_1,s')$. All are samplable. For each transition
$s\to s'$, we can
sample as many states $s_1,s_2$ as we want.

TODO rewrite gradient with current $\zeta$ and previous $\zeta_0$, with
all correct signs and $\gamma$ factors... and collect into a proposition
about $\partial \ell$.

TODO might be less easy to use in policy gradient: because of the
expression for $V$ we have contributions $R(s')$.
